Hydraulically-vulnerable trees survive on deep-water access during droughts in a tropical forest

### 2021-05-07 09.0_manuscript.rmd
rephrased a sentence with evergreen species

- Deep-water access is arguably the most effective, but under-studied mechanism that plants employ to survive during drought. Vulnerability to embolism and hydraulic safety margins can predict mortality risk at given levels of dehydration, but deep-water access may delay plant dehydration. Here, we tested the role of deep-water access in enabling survival within a diverse tropical forest community in Panama using a novel data-model approach.
- We inversely estimated the effective rooting depth (ERD, as the average depth of water extraction), for 29 canopy species by linking diameter growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and leaf hydraulic vulnerability curves. We validated ERD estimates against existing isotopic data of potential water-access depths.
- Across species, deeper ERD was associated with higher maximum stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates during extreme droughts over 35 years (1981-2015), especially in evergreen species. Species exposure to water-stress declined with deeper ERD indicating that trees compensate for water-stress related mortality risk through deep-water access.
- The role of deep-water access in mitigating mortality of hydraulically vulnerable trees has important implications for our predictive understanding of forest dynamics under current and future climates.
\
Keywords: rooting depths, deep-water access, hydraulic vulnerability and safety margins, drought-induced mortality, drought tolerance, tropical forest, safety-efficiency trade-off, hydrological droughts

>>>>

- Deep-water access is arguably the most effective, but under-studied mechanism that plants employ to survive during drought. Vulnerability to embolism and hydraulic safety margins can predict mortality risk at given levels of dehydration, but deep-water access may delay plant dehydration. Here, we tested the role of deep-water access in enabling survival within a diverse tropical forest community in Panama using a novel data-model approach.
- We inversely estimated the effective rooting depth (ERD, as the average depth of water extraction), for 29 canopy species by linking diameter growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and leaf hydraulic vulnerability curves. We validated ERD estimates against existing isotopic data of potential water-access depths.
- Across species, deeper ERD was associated with higher maximum stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates, especially in evergreen species, during extreme droughts over 35 years (1981-2015). Species exposure to water-stress declined with deeper ERD indicating that trees compensate for water-stress related mortality risk through deep-water access.
- The role of deep-water access in mitigating mortality of hydraulically vulnerable trees has important implications for our predictive understanding of forest dynamics under current and future climates.
\
Keywords: rooting depths, deep-water access, hydraulic vulnerability and safety margins, drought-induced mortality, drought tolerance, tropical forest, safety-efficiency trade-off, hydrological droughts
